---
title: '8\. Worksheet: Among Site (Beta) Diversity -- Part 1'
author: "Madison Stoltz; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this worksheet, we move beyond the investigation of within-site $\alpha$-diversity.
We will explore $\beta$-diversity, which is defined as the diversity that occurs among sites.
This requires that we examine the compositional similarity of assemblages that vary in space or time.

After completing this exercise you will know how to:

1. formally quantify $\beta$-diversity
2. visualize $\beta$-diversity with heatmaps, cluster analysis, and ordination
3. test hypotheses about $\beta$-diversity using multivariate statistics

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) with your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercises.
4. Answer questions in the  worksheet.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme). 
5. Before you leave the classroom today, it is *imperative* that you **push** this file to your GitHub repo, at whatever stage you are. Ths will enable you to pull your work onto your own computer.
6. When you have completed the worksheet, **Knit** the text and code into a single PDF file by pressing the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your '6.BetaDiversity' folder.
7. After Knitting, please submit the worksheet by making a **push** to your GitHub repo and then create a **pull request** via GitHub.
Your pull request should include this file 
(**6.BetaDiversity_1_Worksheet.Rmd**) with all code blocks filled out and questions answered) and the PDF output of `Knitr`  
(**6.BetaDiversity_1_Worksheet.pdf**).

The completed exercise is due on **Wednesday, February 1^st^, 2023 before 12:00 PM (noon)**.

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your "*/6.BetaDiversity*" folder, and
4. load the `vegan` R package (be sure to install if needed).  

```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/QB2023_Stoltz/2.Worksheets/6.BetaDiversity")
```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, do the following:  

1. load the `doubs` dataset from the `ade4` package, and
2. explore the structure of the dataset.

```{r}
# note, please do not print the dataset when submitting
package.list <- c('vegan', 'ade4', 'viridis', 'gplots', 'BiodiversityR', 'indicspecies')
for (package in package.list) {
  if (!require(package, character.only=TRUE, quietly=TRUE)) {
    install.packages(package)
    library(package, character.only=TRUE)
  }
}

data(doubs)
str(doubs, max.level=1)
head(doubs$env)
```

***Question 1***:  Describe some of the attributes of the `doubs` dataset.  

a.  How many objects are in `doubs`?
b.  How many fish species are there in the `doubs` dataset?
c.  How many sites are in the `doubs` dataset?

> ***Answer 1a***:  4
> ***Answer 1b***:  27
> ***Answer 1c***:  30

### Visualizing the Doubs River Dataset

***Question 2***: Answer the following questions based on the spatial patterns of richness (i.e., $\alpha$-diversity) and Brown Trout (*Salmo trutta*) abundance in the Doubs River.  

a.  How does fish richness vary along the sampled reach of the Doubs River?
b.  How does Brown Trout (*Salmo trutta*) abundance vary along the sampled reach of the Doubs River?
c.  What do these patterns say about the limitations of using richness when examining patterns of biodiversity?

> ***Answer 2a***:  As the coordinates go from downstream to upstream, the species richness decreases. 
> ***Answer 2b***:  For the Brown Trout, the opposite occurs. As the coordinates go from downstream to upstream, the abundance increases.
> ***Answer 2c***:  Richness does not show distributions of abundance. 

## 3) QUANTIFYING BETA-DIVERSITY

In the R code chunk below, do the following:  

1. write a function (`beta.w()`) to calculate Whittaker's $\beta$-diversity (i.e., $\beta_{w}$) that accepts a site-by-species matrix with optional arguments to specify pairwise turnover between two sites, and
2. use this function to analyze various aspects of $\beta$-diversity in the Doubs River.  

```{r}
beta.w <- function(site.by.species = ""){
  SbyS.pa <- decostand(site.by.species, method = "pa")
  S <- ncol(SbyS.pa[,which(colSums(SbyS.pa)>0)])
  a.bar <- mean(specnumber(SbyS.pa))
  b.w <- round(S/a.bar, 3)
  return(b.w)
}

beta.w <- function(site.by.species = "", sitenum1 = "", sitenum2 = "",
  pairwise = FALSE){
  if (pairwise == TRUE){
    if (sitenum1 == "" | sitenum2 == "") {
      print("Error:please specificy site to compare")
      return(NA)}
    
    site1 = site.by.species[sitenum1,]
    site2 = site.by.species[sitenum2,]
    site1 = subset(site1, select=site1>0)
    site2 = subset(site2, select=site2>0)
    gamma=union(colnames(site1),colnames(site2))
    s = length(gamma)
    a.bar=mean(c(specnumber(site1), specnumber(site2)))
    b.w = round(s/a.bar - 1,3)
    return(b.w)
  }
  else{
    SbyS.pa <- decostand(site.by.species, method = "pa")
    S <- ncol(SbyS.pa[,which(colSums(SbyS.pa)>0)])
    a.bar <- mean(specnumer(SbyS.pa))
    b.w <- round(S/a.bar, 3)
    return(b.w)
  }
}
```

***Question 3***: Using your `beta.w()` function above, answer the following questions:

a.  Describe how local richness ($\alpha$) and turnover ($\beta$) contribute to regional ($\gamma$) fish diversity in the Doubs.
b.  Is the fish assemblage at site 1 more similar to the one at site 2 or site 10?
c.  Using your understanding of the equation $\beta_{w} = \gamma/\alpha$, how would your interpretation of $\beta$ change if we instead defined beta additively (i.e., $\beta = \gamma - \alpha$)?

> ***Answer 3a***:  Local richness and turnover are responsible for defining regional fish diveristy. Changing either richness or turnover will ultimately change regional. 
> ***Answer 3b***:  The fish assemblage at site 1 is more similar to the one at site 10.
> ***Answer 3c***:  If I imagine b=g/a is 1=1/1, then b=g-a would change to 1=1-0. I think a way to interpret this change would be more individuality to gamma and alpha, and therefore less of a relationship between to the two.

### The Resemblance Matrix
In order to quantify $\beta$-diversity for more than two samples, we need to introduce a new primary ecological data structure: the **Resemblance Matrix**. 

***Question 4***: How do incidence- and abundance-based metrics differ in their treatment of rare species?

> ***Answer 4***: Incidence-based metrics consider rare species more heavily than abundance-based metrics. 

In the R code chunk below, do the following:  

1. make a new object, `fish`, containing the fish abundance data for the Doubs River, 
2. remove any sites where no fish were observed (i.e., rows with sum of zero),
3. construct a resemblance matrix based on Sørensen's Similarity ("fish.ds"), and
4. construct a resemblance matrix based on Bray-Curtis Distance ("fish.db").  

```{r}
fish <- doubs$fish
fish <- fish[-8, ]

#calculate the fish who like to dj in their free time
fish.dj <- vegdist(fish, method="jaccard", binary=TRUE)
fish.db <- vegdist(fish, method="bray")
fish.ds <- vegdist(fish, method="bray", binary=TRUE)

fish.db

fish.db <- vegdist(fish, method="bray", upper=TRUE, diag=TRUE)
```

***Question 5***: Using the distance matrices from above, answer the following questions:

a. Does the resemblance matrix (`fish.db`) represent similarity or dissimilarity? What information in the resemblance matrix led you to arrive at your answer? 
b. Compare the resemblance matrices (`fish.db` or `fish.ds`) you just created. How does the choice of the Sørensen or Bray-Curtis distance influence your interpretation of site (dis)similarity?

> ***Answer 5a***: The fish.db resemblance matrix represents similarity, becuase the parts missing are similarity between one sample (full similarity).
> ***Answer 5b***: Bray-Cutris includes '1' and Sorensen only has numbers less than 1. Therefore, their interpretation of similarity is different. 

## 4) VISUALIZING BETA-DIVERSITY

### A. Heatmaps

In the R code chunk below, do the following:  

1. define a color palette,
2. define the order of sites in the Doubs River, and
3. use the `levelplot()` function to create a heatmap of fish abundances in the Doubs River.

```{r}
order <- rev(attr(fish.db, "Labels"))
levelplot(as.matrix(fish.db)[,order], aspect="iso", col.regions=inferno,
          xlab= "Doubs Site", ylab = "Doubs Site", scales=list(cex=0.5),
          main = "Bray-Curtis Distance")
```

### B.  Cluster Analysis

In the R code chunk below, do the following:  

1. perform a cluster analysis using Ward's Clustering, and
2. plot your cluster analysis (use either `hclust` or `heatmap.2`). 

```{r} 
fish.ward <- hclust(fish.db, method="ward.D2")
par(mar=c(1,5,2,2)+0.1)
plot(fish.ward,main= "Doubs River Fish: Ward's Clustering",
     ylab="Squared Brau-Curtis Distance")

gplots::heatmap.2(as.matrix(fish),
                 distfun= function(x) vegdist(x, method="bray"),
                 hclustfun = function(x) hclust(x, method="ward.D2"),
                 col = viridis, trace="none", density.info="none")
```

***Question 6***: Based on cluster analyses and the introductory plots that we generated after loading the data, develop an ecological hypothesis for fish diversity the `doubs` data set? 

> ***Answer 6***: 

### C.  Ordination

### Principal Coordinates Analysis (PCoA)

In the R code chunk below, do the following:  

1. perform a Principal Coordinates Analysis to visualize beta-diversity
2. calculate the variation explained by the first three axes in your ordination
3. plot the PCoA ordination,
4. label the sites as points using the Doubs River site number, and
5. identify influential species and add species coordinates to PCoA plot.

```{r}
fish.pcoa <- cmdscale(fish.db, eig=TRUE, k=3)

explainvar1 <- round(fish.pcoa$eig[1]/sum(fish.pcoa$eig), 3) *100
explainvar2 <- round(fish.pcoa$eig[2]/sum(fish.pcoa$eig), 3) *100
explainvar3 <- round(fish.pcoa$eig[3]/sum(fish.pcoa$eig), 3) *100
sum.eig <- sum(explainvar1, explainvar2, explainvar3)

par(mar = c(5,5,1,2)+0.1)

plot(fish.pcoa$points[,1], fish.pcoa$points[,2], ylim=c(-0.2, 0.7),
     xlab = paste("PCoA 1 (", explainvar1, "%)", sep=""),
     ylab = paste("PCoA 2 (", explainvar2, "%)", sep=""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5,
     cex.axis = 1.2, axes=FALSE)

axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h=0, v=0, lty=3)
box(lwd=2)

points(fish.pcoa$points[,1], fish.pcoa$points[,2],
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(fish.pcoa$points[,1], fish.pcoa$points[,2],
     labels = row.names(fish.pcoa$points))

fishREL <- fish
  for(i in 1:nrow(fish)){
    fishREL[i,]=fish[i,]/sum(fish[i,])
  }

fish.pcoa <- add.spec.scores(fish.pcoa, fishREL, method="pcoa.scores")
text(fish.pcoa$cproj[,1], fish.pcoa$cproj[,2],
     labels = row.names(fish.pcoa$cproj), col = "black")
```

In the R code chunk below, do the following:  

1. identify influential species based on correlations along each PCoA axis (use a cutoff of 0.70), and
2. use a permutation test (999 permutations) to test the correlations of each species along each axis.

```{r}
spe.corr <- add.spec.scores(fish.pcoa, fishREL, method = "cor.scores")$cproj
corrcut <- 0.7
imp.spp <- spe.corr[abs(spe.corr[,1])>= corrcut | abs(spe.corr[,2]) >= corrcut, ]

fit <- envfit(fish.pcoa, fishREL, perm=999)
```

***Question 7***: Address the following questions about the ordination results of the `doubs` data set:

a.  Describe the grouping of sites in the Doubs River based on fish community composition.
b.  Generate a hypothesis about which fish species are potential indicators of river quality.

> ***Answer 7a***: There is a large overlap at x-axis at 0, but also a general spread througout the x-axis. For the y-axis, most groups are less than .2. 
> ***Answer 7b***: If fish species are more abundant in an area, then the area they are in is of higher river quality compared to areas that have low species abundance. 


## SYNTHESIS

Load the dataset from that you and your partner are using for the team project. 
Use one of the tools introduced in the beta diversity module to visualize your data. 
Describe any interesting patterns and identify a hypothesis is relevant to the principles of biodiversity. 


```{r}
library(readr)
Data_DRYAD_Seabirds <- read_csv("~/Desktop/Data_DRYAD_Seabirds.csv")
```

> Synthesis Answer: I am having a difficult time trying to run one of the tools on my data. I contacted my partner but they have not responded. I think an issue may stem from needing to change the data in a way that r can read it better. I plan to ask about this issue on Friday before class.